---
title: Gapminder analysis
author: Ben Yang
output:
    html_document:
        theme: yeti
        highlight: pygments
        code_folding: hide
        toc: true
        toc_float: true
        number_sections: true
        fig_width: 7
        fig_height: 6
        df_print: paged
---

# Guided section

```{r message=F}
library(tidyverse)
library(plotly)
```

## Read data

We read the data given in a CSV file, discard the first column of indices, and
rename some variables for convenience.

```{r results='hide', message=F, warning=F}
data <- read_csv("./gapminder_clean.csv") %>%
    select(-1) %>%
    rename(
       co2em = `CO2 emissions (metric tons per capita)`,
       popden = `Population density (people per sq. km of land area)`,
       lifeExp = `Life expectancy at birth, total (years)`,
       energy = `Energy use (kg of oil equivalent per capita)`,
       imports = `Imports of goods and services (% of GDP)`,
    )
```

## Scatter plot of CO2 emissions and GDP in 1962

We take rows of data from 1962 and look at their GDP and CO2 emissions. We
throw away rows that have a NA value for either variable. Below is a scatter
plot of this filtered data. Each dot represents a country.

```{r}
data1962 <- data %>%
    filter(Year == 1962) %>%
    select(gdpPercap, co2em) %>%
    drop_na()
ggplot(data = data1962) +
    geom_point(mapping = aes(
        x = gdpPercap,
        y = co2em)) +
    labs(title = "GDP vs CO2 emissions per capita in 1962",
        x = "GDP per capita",
        y = "CO2 emissions per capita (metric tons)")
```

## Pearson correlation of CO2 emissions and GDP in 1962

```{r}
cor.test(data1962$gdpPercap, data1962$co2em)
```

Our p-value is less than 0.001. The correlation coefficient suggests that the
two variables are positively associated.

## Year of strongest correlation between CO2 emissions and GDP

We group our data by year. For each group, we find the correlation. Lastly, we
find which year has the maximum correlation.

```{r results='hide', message=F}
corrs <- data %>%
    group_by(Year) %>%
    select(Year, gdpPercap, co2em) %>%
    drop_na() %>%
    summarise(correlation = cor(gdpPercap, co2em))
maxi <- lapply(corrs, max)
```

The strongest correlation is __`r maxi$correlation`__ in the year __`r maxi$Year`__.

## Interactive scatter plot of CO2 emissions and GDP in the year with the highest correlation

We filter our data using the year we found in the previous section.

```{r, warning=F}
max_em_year_data <- data %>%
    filter(Year == maxi$Year) %>%
    select(gdpPercap, co2em, pop, continent, `Country Name`) %>%
    drop_na()
fig <- ggplot(data = max_em_year_data) +
    geom_point(aes(
        x = gdpPercap,
        y = co2em,
        size = pop,
        color = continent,
        text = paste("Country: ", `Country Name`,
             "\nGDP: ", gdpPercap,
             "\nCO2 emissions: ", co2em))) +
    labs(title = str_glue("GDP vs CO2 emissions per capita in ", maxi$Year),
        x = "GDP per capita",
        y = "CO2 emissions per capita (metric tons)")
ggplotly(fig, tooltip = "text")
```

# Open section

For sections requiring a statistics test, I use the standard `alpha = 0.95`.

## Relationship between continent and energy use

Let's get a general idea of what we're looking at. We filter our data down
to the continents and energy usages.

```{r}
data_energy <- data %>%
    select(continent, energy) %>%
    drop_na()
ggplot(data = data_energy) +
    facet_wrap(continent ~ .) +
    geom_histogram(
        mapping = aes(x = energy),
        position = "identity",
        binwidth = 500,
    ) +
    labs(title = "Histograms of energy usage per continent 1962-2007",
        x = "Energy use (kg of oil equivalent per capita)",
        y = "Count"
    )
```

All of them except Oceania look right-skewed. Oceania's plot is quite small, so
let's view it on it's own.

```{r}
ggplot(data = data_energy %>% filter(continent == "Oceania")) +
    geom_histogram(
        mapping = aes(x = energy),
        position = "identity",
        binwidth = 200,
    ) +
    labs(title = "Histogram of energy usage in Oceania 1962-2007",
        x = "Energy use (kg of oil equivalent per capita)",
        y = "Count"
    )
```

We don't have a lot of data for Oceania, so it'll look blocky. I can't say much
about it. The next question is _can we perform ANOVA_ and statistically
determine relationships in the data. First, we check if the data satisfy the
assumptions required to use ANOVA.

### Normality

Let's observe how normal the data are with Q-Q plots.

```{r}
ggplot(data_energy, mapping = aes(sample = energy)) +
    facet_wrap(continent ~ ., nrow = 2) +
    stat_qq() +
    stat_qq_line() +
    theme(aspect.ratio = 1) +
    labs(title = "Q-Q plots of energy use per continent 1962-2007",
        x = "Theoretical quantiles",
        y = "Energy use (kg of oil equivalent per capita) quantiles"
    )
```

Africa, Oceania and Europe look approximately normal, but the others do not.
Let's apply a data transformation. We get an empty tibble when filtering for
energy values less than or equal to zero. So, it is safe to apply the logarithm
to the energy usage data.

```{r}
data_energy %>%
    filter(energy <= 0)
```

Let's try the data transformation.

```{r}
data_energy_t <- data_energy %>%
    mutate(energy_transformed = log(energy))
ggplot(data_energy_t, mapping = aes(sample = energy_transformed)) +
    facet_wrap(continent ~ ., nrow = 2) +
    stat_qq() +
    stat_qq_line() +
    theme(aspect.ratio = 1) +
    labs(title = "Q-Q plots of log energy use per continent 1962-2007",
        x = "Theoretical quantiles",
        y = "Log of energy use (kg of oil equivalent per capita) quantiles"
    )
```

Oceania remains normal. The other continents' data look more normal now. They
still are not perfect, but it's okay since they have relatively large sample
sizes, as the table below illustrates.

```{r}
data_energy_t %>%
    group_by(continent) %>%
    summarise(count = n())
```

We continue, assuming the data are approximately normal.

### Independence

The data come from different continents and are at least 5 years apart. Due to
this, assume the data are independent and identically distributed.

### Equality of variances

How do the variances look?

```{r}
ggplot(data_energy_t) +
    geom_boxplot(mapping = aes(energy_transformed, continent)) +
    labs(title = "Variances of log of energy usage per continent",
        x = "Log of energy usage (kg of oil equivalent per capita)",
        y = "Continent")
```

The variances of the transformed data look roughly the same. Oceania stands out
as having a smaller variance than the others. Asia looks like it has a larger
variance. However, they don't look too equal. We proceed with caution.

### ANOVA

```{r}
test_results <- aov(energy_transformed ~ continent, data = data_energy_t)
summary.aov(test_results)
```

The p-value is less than 0.001. In our assumptions, we saw that the transformed
data were not exactly normal, and their variances were not exactly equal.
Taking that into account, I decide to reject the null hypothesis. That is, we
reject that the log of the group means are equal. Since the logarithm is a
monotone increasing function, we can reject that the group means are equal.

### Post hoc testing via Tukey's

We run Tukey's test to determine differences between pairs of continents.

```{r}
TukeyHSD(test_results)
```

For every pair except for (Asia, Americas) and (Oceania, Europe), the p-value
is less than 0.0001, so for those, we reject the null hypothesis that the
logarithm of the energy usage means are the same. For the exceptional pairs, we
retain the null hypothesis.

## Difference of imports between Europe and Asia

We filter our data to imports of goods and services in Europe and Asia. Let's
take a look at a histogram.

```{r}
data_imports <- data %>%
    filter(Year >= 1990 & continent %in% c("Europe", "Asia")) %>%
    select(continent, imports) %>%
    drop_na()
ggplot(data = data_imports) +
    geom_histogram(
        mapping = aes(x = imports, fill = continent),
        position = "identity",
        alpha = 0.5,
        binwidth = 10,
    ) +
    labs(title = "Histograms of imports in Europe and Asia",
        x = "Imports of goods and services (% of GDP)",
        y = "Count"
    )
```

The peaks seem to line up, but Europe as a bunch more data points near the
peaks, compared to Asia. At a glance, I would guess that the two continents
have roughly the same imports of goods and services, measured in percent of
GDP. Can we perform an unpaired t-test to check our guess? Let's check the
assumptions.

### Normality

We have histograms above, but I find Q-Q plots easier to look at.

```{r}
ggplot(data_imports, mapping = aes(sample = imports)) +
    facet_wrap(continent ~ ., nrow = 1) +
    stat_qq() +
    stat_qq_line() +
    theme(aspect.ratio = 1) +
    labs(title = "Q-Q plots of imports in Europe and Asia",
        x = "Theoretical quantiles",
        y = "Imports of goods and services (% of GDP) quantiles"
    )
```

It looks kind of normal, but can we do better with a data transformation?

```{r}
data_imports_t <- data_imports %>%
    mutate(imports_t = sqrt(imports))
ggplot(data_imports_t, mapping = aes(sample = imports_t)) +
    facet_wrap(continent ~ ., nrow = 1) +
    stat_qq() +
    stat_qq_line() +
    theme(aspect.ratio = 1) +
    labs(title = "Q-Q plots of log imports in Europe and Asia",
        x = "Theoretical quantiles",
        y = "Log of imports of goods and services (% of GDP) quantiles"
    )
```

The transformation helps a bit, but the data at the ends still do not fit the
lines well. Let's proceed skeptically, assuming that the data are approximately
normal.

### Independence

Europe and Asia are completely different regions of land. Assume each datum
comes from only one of the continents.

 We use the time argument again. The data are taken 5 years apart, so for each
 group of data, assume that the data are independent and identically
 distributed.

### Outliers

```{r}
ggplot(data_imports_t) +
    geom_boxplot(mapping = aes(imports_t, continent)) +
    labs(title = "Outliers in the square root of imports in Europe vs Asia",
        x = "Square root of imports of goods and services (percent of GDP)",
        y = "Continent")
```

We have our eyes on Welch's t-test, which is robust against unequal variances. For a normal unpaired t-test, we'd still be fine because the sample sizes are roughly equivalent.

```{r}
data_imports_t %>%
    group_by(continent) %>%
    summarise(count = n())
```

We see outliers on the Asia data, which may affect our results negatively.
We continue anyways, acknowledging these shortcomings.

### Welch's two sample t-test

```{r}
x <- data_imports_t %>% filter(continent == "Europe")
y <- data_imports_t %>% filter(continent == "Asia")
t.test(x$imports, y$imports)
```

Our p-value is 0.1776. I'm skeptical because we did not satisfy some of our
parametric assumptions as much as I wanted. We run the nonparametric
Mann-Whitney test for assurance.

### Mann-Whitney U test

```{r}
wilcox.test(x$imports, y$imports)
```

Our p-value here is 0.7867. Both tests indicate that we can retain the null
hypothesis. However the null hypothesis of the parametric and nonparametric
tests are different. In this case, I retain the null hypothesis of the
parametric test because the assumptions were not completely wrong, and the
statement is stronger.

That is, we retain the null hypothesis that square root of
import means are the same. Because the square root function is monotone
increasing, this is the same as saying the import means between Europe and Asia
are the same.

## Countries with the highest population density in 1962-2007

For each available year, we list the 10 countries with the highest population
densities.

```{r, warning=F}
data_popden <- data %>%
    group_by(Year) %>%
    select(Year, `Country Name`, popden) %>%
    drop_na() %>%
    slice_max(n = 10, order_by = popden)
fig <- ggplot(data = data_popden) +
    geom_line(mapping = aes(
        x = Year,
        y = popden,
        group = `Country Name`,
        color = `Country Name`,
        text = paste("Country: ", `Country Name`,
             "\nDensity: ", popden,
             "\nYear: ", Year))
    ) +
    labs(title = "10 most population dense countries per year 1962-2007",
        x = "Year",
        y = "Population density (people per sq. km of land area)"
    )
ggplotly(fig, tooltip = "text")
```

The countries with the highest population densities within 1962 and 2007 are
__Macao SAR, China__ and __Monaco__.

## Country with the greatest increase in life expectancy at birth since 1962

We group the data by country. In each group, we create a new column that tracks
the change in life expectancy since 1962. The following figure shows this data for
some of the top growers.

```{r}
data_life_exp <- data %>%
    select(`Country Name`, Year, lifeExp) %>%
    drop_na() %>%
    group_by(`Country Name`) %>%
    filter(any(Year == 1962)) %>%
    mutate(lifeExpSince1962 = lifeExp - lifeExp[Year == 1962]) %>%
    ungroup()

countries_highest <- data_life_exp %>%
    group_by(`Country Name`) %>%
    mutate(
        lifeExpTotalChange = lifeExp[Year == 2007] - lifeExp[Year == 1962]) %>%
    summarise(lifeExpChange = max(lifeExpTotalChange)) %>%
    arrange(desc(lifeExpChange))

cutoff_life_exp_change <- min(head(countries_highest, n = 10)$lifeExpChange)

data_life_exp_top <- data_life_exp %>%
    group_by(`Country Name`) %>%
    filter(
        lifeExp[Year == 2007] - lifeExp[Year == 1962] >= cutoff_life_exp_change
    )

fig <- ggplot(data = data_life_exp_top) +
    geom_line(mapping = aes(
        x = Year,
        y = lifeExpSince1962,
        color = `Country Name`)
    ) +
    labs(title = "Increase in life expectancy of top 10 countries 1962-2007",
        x = "Year",
        y = "Change in life expectancy at birth since 1962 (years)")
ggplotly(fig)
```

The __Maldives__ had the highest increase in life expectancy at birth from 1962 to 2007.
